\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-09-21 Manuscript/manuscript.tex
one more round on intro

Devon Allen, a University of Oregon alumnus, was a strong 
contender for the 2022 World Track and Field Championships 
in the 110-meter hurdle event. After running the event in 
12.84 seconds in June 2022, just 0.04 seconds short of the 
world record \citep{wa2022preview}, and placing third at 
the U.S. Track and Field Championships, Allen was well-prepared 
for the World Championships in August, also held in Eugene, 
Oregon. He advanced through the preliminary heats and semifinals 
with reaction times of 0.123 and 0.101 seconds, respectively. 
However, in the final heat, Allen was disqualified for a 
reaction time of 0.099 seconds, just 0.001 seconds faster 
than the 0.1-second threshold set by the International 
Association of Athletics Federations (IAAF). This decision 
was met with significant public outcry \citep{nbc2022devon}.

Allen's disqualification has been discussed extensively in online 
communities such as \url{www.LetsRun.com}, a website that functions 
as both a message board and a news outlet. The message board, 
similar to Reddit, is very active during major running events, 
including the Olympics and the World Track and Field Championships. 
The news section of the website featured articles such as 
\citep{johnson2022data} and \citep{johnson2022was} by Robert Johnson, 
the site’s founder. Johnson argued that there may have been an 
error with the timing equipment at the 2022 Championships. He 
cited graphs, descriptive statistics like median reaction times, 
and compared the reaction times at the 2022 Championships to 
those from other competitions, including the U.S. Championships 
and previous World Championships. However, these comparisons 
lacked formal statistical analysis. Johnson estimated a 99.95\% 
chance that Allen was wrongfully disqualified, based on his claim 
that the probability of the fastest reaction times from the men's 
100 meter, men's 110 meter hurdle, and women's 100 meter all 
occurring in the same year was approximately $(1/13)^3$ 
\citep{johnson2022was}.

The first objective of this paper is to determine whether athletes’ 
reaction times at the 2022 World Track and Field Championships 
were significantly different from other competitions. Our analysis 
compares reaction times for athletes who competed in multiple events, 
using a matched-pairs design. By comparing the same athletes' reaction
times across different competitions, we can assess whether the 2022 
reaction times differed significantly. While this investigation was 
motivated by Allen's disqualification, it also addresses broader concerns 
about potential variation in timing equipment across different venues
in short-distance dash or hurdle.

Given the observed variation in reaction times across venues, the 
question arises: is the 0.1-second disqualification threshold truly 
justified? Studies have presented evidence supporting both sides of 
the debate. \citet{pain2007sprint} examined reaction times in nine 
male athletes and found that the fastest athlete had an average 
reaction time of 0.087 seconds, with a standard deviation of 0.004 
seconds. Similarly, \citet{komi2009iaaf}, in a study commissioned by 
the IAAF, recommended lowering the disqualification threshold to 0.08 
or 0.085 seconds, arguing that athletes are capable of faster reaction 
times than the current threshold allows. On the other hand, some 
researchers have argued for increasing the reaction time threshold. 
\citet{brosnan2017effects} suggested raising it to 0.115 seconds based 
on data from European and World Championships, while a study of the 
2008 Olympics supported the conclusion that athletes cannot reliably 
react within 0.1 seconds \citep{lipps2011implications}. Despite these 
findings, the IAAF has not revised the threshold since its implementation 
in 1999. It is important to note that \citet{brosnan2017effects} and 
\citet{lipps2011implications} relied on retrospective event data, 
whereas \citet{pain2007sprint} and \citet{komi2009iaaf} conducted 
controlled experiments to assess reaction time capabilities.

The second objective of this paper is to model reaction times for 
non-disqualified athletes from the 1999 World Championships onward. 
We employ a gamma linear mixed-effects model with two random effects: 
one for the venue (year) and another for the heat nested within each venue.
This model allows us to estimate the probability of an athlete's 
reaction time falling below a specific threshold, such as 0.1 seconds. 
By examining reaction times across years and heats, we aim to assess 
the consistency of reaction times and whether the current 
disqualification threshold is statistically justified. This modeling 
approach provides a more detailed understanding of the factors that 
influence reaction times, offering insights into whether the 0.1-second 
threshold should be reconsidered for future competitions.

>>>>

Devon Allen’s highly anticipated performance at the 2022 World 
Track and Field Championships in Eugene, Oregon, ended in 
controversy when he was disqualified for a reaction time of 
0.099 seconds, just 0.001 seconds below the allowable threshold.
Allen, a University of Oregon alumnus, 
had recently run a time of 12.84 seconds in the 110-meter hurdle 
event, just 0.04 seconds short of the world record. After placing 
third at the U.S. Track and Field Championships, he advanced 
through the preliminary heats and semifinals at the World 
Championships, with reaction times of 0.123 and 0.101 seconds, 
respectively. However, in the final heat, competing in front of his 
home audience, Allen’s reaction time was just 0.001 seconds faster 
than the 0.1-second threshold set by the International Association 
of Athletics Federations (IAAF), resulting in his disqualification, a 
decision that was met with widespread public outcry.

The disqualification of Allen was widely discussed in online 
communities, with some questioning the accuracy of the timing 
equipment. One of the most active platforms was \url{www.LetsRun.com}, 
a website that functions as both a message board and a news outlet. 
The message board, similar to Reddit, is particularly active during 
major running events, including the Olympics and the World Track 
and Field Championships. The news section of the website featured 
articles such as \citep{johnson2022data} and \citep{johnson2022was} 
by Robert Johnson, the site’s founder. Johnson argued that there may 
have been an error with the timing equipment at the 2022 Championships. 
He cited graphs and descriptive statistics, like median reaction 
times, comparing the reaction times at the 2022 Championships to 
those from other competitions, including the U.S. Championships and 
previous World Championships. However, these comparisons lacked formal 
statistical analysis.

The first objective of this paper is to investigate whether reaction
times at the 2022 World Championships were significantly different
from other competitions, focusing specifically on athletes who
competed in multiple events. By employing a matched-pairs design, we
compare reaction times from the same athletes across different
championships, including the 2022, 2019, and 2023 World Championships,
as well as the 2022 national-level competitions. This approach allows
us to isolate the effect of the competition year on reaction times
while controlling for individual athlete performance. Our analysis
provide a more formal statistical evaluation of the disqualification 
threshold, building upon the informal comparisons previously made 
by others.

There is an ongoing debate over the appropriateness of the 0.1-second
threshold for disqualification, with past studies suggesting both
lower and higher limits. \citet{pain2007sprint} examined 
reaction times in nine male athletes and found that the fastest athlete 
had an average reaction time of 0.087 seconds, with a standard deviation 
of 0.004 seconds. Similarly, \citet{komi2009iaaf}, in a study commissioned 
by the IAAF, recommended lowering the disqualification threshold to 0.08 
or 0.085 seconds, arguing that athletes are capable of faster reaction 
times than the current threshold allows. On the other hand, some 
researchers have argued for increasing the threshold.
\citet{brosnan2017effects} suggested raising it to 0.115 seconds based
on data from European and World Championships, while a study of the
2008 Olympics supported the conclusion that athletes cannot reliably
react within 0.1 seconds \citep{lipps2011implications}. Interestingly,
\citet{brosnan2017effects} and \citet{lipps2011implications} relied on
retrospective event data, whereas \citet{pain2007sprint} and
\citet{komi2009iaaf} conducted controlled experiments to assess
reaction time capabilities. The IAAF has not revised the threshold
since its implementation in 1999.

The second objective of this paper is to further investigate the
appropriateness of the 0.1-second threshold by modeling reaction times
from athletes who competed in world championships from 1999 onward. We
employ a gamma linear mixed-effects model with two random effects: one
for the venue (year) and another for the heat nested within each
venue. This model allows us to estimate the probability of an
athlete’s reaction time falling below a specific threshold, such as
0.1 seconds. By examining reaction times across different years and
heats, we can assess the consistency of reaction times and whether the
current disqualification threshold is supported by statistical
evidence.
